Sequential rank agreement methods for comparison of ranked lists

### 2016-03-31 latex/paper.Rnw
Starting on the asymptotic

@@

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p)),
\end{equation}
for a non-negative real-valued distance function $f.$ Throughout this
paper we will use the sample standard error as our function $f$ and
hence use
$$A(X_p) = \sqrt{\frac{\sum_{i=1}^L (R_i(X_p) - \bar{R}(X_p))^2}{L-1}},
$$
where $\bar{R}(X_p)$ is the average rank assigned to item $X_p$, but
other possibilities will be discussed below. For each item, the sample
standard error has an interpretation as the average distance of the
individual rankings from the average ranking over all the lists, and
it corresponds to the same measure that is used in method comparison
studies to compute the limits of agreement \citep{alt:bland:1983}.

For an integer $1\le d\le P$ we define the unique set of items found
by merging the first $d$ elements of each of the $L$ lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \},
\end{equation}
which is exemplified in Panel (c) of Table
\ref{tab:example}.

The \emph{sequential rank agreement} is the pooled standard
deviation of the items found in the set $S_d$:
\begin{equation}
\textrm{sra}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
where $|S_d|$ refers to the cardinality of the set $S_d$. Values of
\textrm{sra} close to zero suggests that the lists agree while larger
values suggest disagreement. Just like in method comparison studies we
prefer smaller limits of agreement because that means that the
differences between the methods is small. If the ranked lists are
identical then the value of sequential rank agreement will be zero for
all depths $d$. The sequential rank agreement can be interpreted as
the average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.

Note that the terms that are part of \eqref{def:sra} are not
independent since every list contains the ranks from 1 to $P$ exactly
once. As a consequence we simply use the pooled standard deviation as
a measure of the average distance between rankings and we refrain from
using traditional distributional results for this measure.

@@

>>>>

\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\arginf}{arginf}

\usepackage[amsthm,thmmarks]{ntheorem}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\newtheorem{example}{Example}
\renewcommand\proofSymbol{\ensuremath{\blacksquare}}

@@

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p))
\end{equation}
for a measurable non-negative real-valued distance function $f$.
For an integer $1\le d\le P$ we define the unique set of items found
by merging the first $d$ elements of each of the $L$ lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.
We then define the \emph{sequential rank agreement} as pooled expected agreement of the items
found in the set $S_d$:
\begin{equation}
\textrm{sra}(d)= \frac{\sum_{\{p \in S_d\}}(L-1)\E[A(X_p)]}{(L-1)|S_d|}\label{def:sra}
\end{equation}
where $|S_d|$ refers to the cardinality of the set $S_d$, and the expectation
is taken with respect to the distribution of the rankings of $X_p$ accross the $L$ lists.
Values of \textrm{sra} close to zero suggests that the lists agree while larger
values suggest disagreement. Just like in method comparison studies we
prefer smaller limits of agreement because that means that the
differences between the methods is small. If the ranked lists are
identical then the value of sequential rank agreement will be zero for
all depths $d$.

%Note that the terms that are part of \eqref{def:sra} are not
%independent since every list contains the ranks from 1 to $P$ exactly
%once. As a consequence we simply use the pooled standard deviation as
%a measure of the average distance between rankings and we refrain from
%using traditional distributional results for this measure.

Throughout the rest of this paper we will use the variance to quantify the
agreement of the rankings and therefore define
\begin{align}
  A(X_p) = (R_i(X_p) -  \E[R_i(X_p)])^2
\end{align}
For each item, the function $A$ has an interpretation as the expected
euclidean distance of the individual rankings from the expected ranking over all the lists, and
it corresponds to the same measure that is used in method comparison studies to compute the
limits of agreement \citep{alt:bland:1983}. The sequential rank agreement then becomes equivalent
to the pooled variance of the items found in $S_d$, and it can be interpreted as
the average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.

Given a sample of $1,\ldots,L$ lists and estimator of estimated sequatial rank agreement is given by
\begin{equation}
\widehat{\textrm{sra}}_L(d)= \frac{\sum_{\{p \in S_d\}}(L-1)\widehat{A_L}(X_p)}{(L-1)|S_d|}\label{def:sra}
\end{equation}
where
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p)
\end{align}
is the sample variance of the rank of item $X_p$. For an easier interpretation we suggest to plot $\widehat{\textrm{sra}}_L(d)^{1/2}$ corresponding to the pooled standard deviation.

@@

\section{Change-point analysis}
In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek, we work with an infinite set of lists
$\mathcal{L}$ and study the asymptotic behaviour for
$L\to\infty$. However, we do not allow the length of the lists to
change with $L$ as in most applications the length of the lists is
fixed [FIXME: for finite $L$ are there only finitely many forskelige
lists (aber dass stoert keinen grossen geist, siger karlsson)]. First,
we define the parameter of interest. For any given threshold value $q$
a change-point $d^\ast(q)$ in the list agreement is defined by
\begin{equation}
  d^\ast(q) = \arginf_{d\in\{1,\dots,P\}}\textrm{sra}(d)>q.
\end{equation}
The value of $q$ could for example be the agreement of randomly
permuted lists [FIXME: there is only one SRA value for infinitely many
random permutations, no skygge!].
The corresponding estimated change-point is similarly defined by
\begin{equation}
\widehat{d^\ast_L}(q) = \arginf_{d\in\{1,\dots,P\}}\widehat{\textrm{sra}}_L(d)>q
\end{equation}

We start by showing that $\widehat{\textrm{sra}}_L(d)$ is consistent for $\textrm{sra}(d)$ at each
$d \in \left\{1,\ldots,P\right\}$ for $L \rightarrow \infty$. The consistency of $\widehat{d^\ast_L}(q)$
then follows by a simple corollary.

\begin{theorem}
hej
\begin{proof}
We start by noting that we can rewrite the sequential rank agreement in equation \ref{def:sra} as
\begin{equation}
\textrm{sra}(d)= \frac 1{\sum_{p=1}^P \sum_{l=1}^L 1\{R_l(X_p)\le d\}}\sum_{p=1}^P(1-\prod_{l=1}^L 1\{R_l(X_p)>d\}) \E[A(X_p)]
\end{equation}

[FIXME: Andreas details the proof with Cauchy-Schwarz.]
[FIXME: Det er lidt kedelig for saadan en fin en som Andreas: Skal theoremet og bevis ikke vaere for en general agreement metric, ikke nodvendigvis den i equation (2), og kan vi charakterisere hvad den skal opfylde?]
\end{proof}
\label{thm:consistency}
\end{theorem}

\begin{corollary}
For fixed $q$ it follows from theorem \ref{thm:consistency}
  $\widehat{d^\ast_L}(q) \overset{P}{\longrightarrow} d^\ast(q)$
\begin{proof}
 hej
\end{proof}
\end{corollary}

### 2016-03-31 latex/paper.tex
Starting on the asymptotic

\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}

\usepackage[amsthm,thmmarks]{ntheorem}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem*{proposition2}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\newtheorem{example}{Example}
\renewcommand\proofSymbol{\ensuremath{\blacksquare}}

@@

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Mar 31 16:50:36 2016
\begin{table}[tb]
\centering
\caption{Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & T & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1995 & 1037 \\ 
  5 & 829 & 766 & 808 & 2124 \\ 
  6 & 2851 & 2670 & 523 & 808 \\ 
  7 & 703 & 2939 & 1042 & 108 \\ 
  8 & 2386 & 2386 & 2489 & 515 \\ 
  9 & 2645 & 1834 & 1665 & 2670 \\ 
  10 & 2002 & 378 & 849 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

\section{Change-point analysis}

In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek, we work with an infinite set of lists
$\mathcal{L}$ and study the asymptotic behaviour for
$L\to\infty$. However, we do not allow the length of the lists to
change with $L$ as in most applications the length of the lists is
fixed [FIXME: for finite $L$ are there only finitely many forskelige
lists (aber dass stoert keinen grossen geist, siger karlsson)]. First,
we define the parameter of interest. For any given threshold value $q$
a change-point $d_0(q)$ in the list agreement is defined by
\begin{equation*}
  d_0=\textrm{arginf}_{d\in\{1,\dots,L\}}\textrm{sra}(d)>q.
\end{equation*}
The value of $q$ could for example be the agreement of randomly
permuted lists [FIXME: there is only one SRA value for infinitely many
random permutations, no skygge!].

The theorem shows consistency of the plug-in estimate of
\begin{equation*}
\hat{d_0} = \textrm{arginf}_{d\in\{1,\dots,L\}}\widehat{\textrm{sra}}(d)>q
\end{equation*}
[FIXME: $\textrm{sra}$ -> $\widehat{\textrm{sra}}$ in rest of article?]

\begin{theorem}
  $\hat{d_0}\overset{L\to\infty}{\longrightarrow}d_0$

\begin{proof}
Proof: We rewrite the square of the sequential rank agreement \eqref{def:sra} as
\begin{equation}
(\textrm{sra}(d))^2= \frac 1{\sum_{p=1}^P \sum_{l=1}^L 1\{R_l(X_p)\le d\}}\sum_{p=1}^P(1-\prod_{l=1}^L 1\{R_l(X_p)>d\}) A(X_p)^2
\end{equation}
[FIXME: Andreas details the proof with Cauchy-Schwarz.]
[FIXME: Det er lidt kedelig for saadan en fin en som Andreas: Skal theoremet og bevis ikke vaere for en general agreement metric, ikke nodvendigvis den i equation (2), og kan vi charakterisere hvad den skal opfylde?]
\end{proof}
\end{theorem}

>>>>

\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\arginf}{arginf}

\usepackage[amsthm,thmmarks]{ntheorem}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\newtheorem{example}{Example}
\renewcommand\proofSymbol{\ensuremath{\blacksquare}}

@@

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Mar 31 17:59:03 2016
\begin{table}[tb]
\centering
\caption{Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & T & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1524 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

\section{Change-point analysis}
In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek, we work with an infinite set of lists
$\mathcal{L}$ and study the asymptotic behaviour for
$L\to\infty$. However, we do not allow the length of the lists to
change with $L$ as in most applications the length of the lists is
fixed [FIXME: for finite $L$ are there only finitely many forskelige
lists (aber dass stoert keinen grossen geist, siger karlsson)]. First,
we define the parameter of interest. For any given threshold value $q$
a change-point $d^\ast(q)$ in the list agreement is defined by
\begin{equation}
  d^\ast(q) = \arginf_{d\in\{1,\dots,P\}}\textrm{sra}(d)>q.
\end{equation}
The value of $q$ could for example be the agreement of randomly
permuted lists [FIXME: there is only one SRA value for infinitely many
random permutations, no skygge!].
The corresponding estimated change-point is similarly defined by
\begin{equation}
\widehat{d^\ast_L}(q) = \arginf_{d\in\{1,\dots,P\}}\widehat{\textrm{sra}}_L(d)>q
\end{equation}

We start by showing that $\widehat{\textrm{sra}}_L(d)$ is consistent for $\textrm{sra}(d)$ at each
$d \in \left\{1,\ldots,P\right\}$ for $L \rightarrow \infty$. The consistency of $\widehat{d^\ast_L}(q)$
then follows by a simple corollary.

\begin{theorem}
hej
\begin{proof}
We start by noting that we can rewrite the sequential rank agreement in equation \ref{def:sra} as
\begin{equation}
\textrm{sra}(d)= \frac 1{\sum_{p=1}^P \sum_{l=1}^L 1\{R_l(X_p)\le d\}}\sum_{p=1}^P(1-\prod_{l=1}^L 1\{R_l(X_p)>d\}) \E[A(X_p)]
\end{equation}

[FIXME: Andreas details the proof with Cauchy-Schwarz.]
[FIXME: Det er lidt kedelig for saadan en fin en som Andreas: Skal theoremet og bevis ikke vaere for en general agreement metric, ikke nodvendigvis den i equation (2), og kan vi charakterisere hvad den skal opfylde?]
\end{proof}
\label{thm:consistency}
\end{theorem}

\begin{corollary}
For fixed $q$ it follows from theorem \ref{thm:consistency}
  $\widehat{d^\ast_L}(q) \overset{P}{\longrightarrow} d^\ast(q)$
\begin{proof}
 hej
\end{proof}
\end{corollary}
